---
title: "Supervised gene module discovery with Grand Forest"
author: "Simon J. Larsen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

This vignette demonstrates how to use Grand Forest to find a gene module associated with survival in non-small cell lung cancer.

## Network preparation

First we need to prepare the gene-gene interaction network. We can obtain a protein-protein interaction network from BioGRID using the [simpIntLists](https://bioconductor.org/packages/release/data/experiment/html/simpIntLists.html) package from Bioconductor.

```{r}
library(data.table)
library(tidyverse)
library(simpIntLists)

data("HumanBioGRIDInteractionEntrezId")

# convert edge lists into two-column data frame
edges <- lapply(HumanBioGRIDInteractionEntrezId, function(x) {
  data.frame(
    source = as.character(x$name),
    target = as.character(x$interactors),
    stringsAsFactors = FALSE
  )
})
edges <- rbindlist(edges)

head(edges)
```

The resulting data frame should contain two columns with gene identifiers as `character` strings. In this example we will be using Entrez Gene IDs, but any other identifier can be used, provided the same type is used in the experimental data as well.

## Expression data preparation

Next we download a gene expression data set from non-small cell lung cancer patients.
The dataset was extracted from [GSE30219](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE30219).
Besides gene expression values, we also obtain survival information for each patient. The `os_time`-column contains the follow-up time for each patient in months, and the `os_event` indicates whether an event has happened (0 = no event, 1 = death). Each remaining column corresponds to a gene.

```{r}
# Download lung cancer expression data set with survival times.
D <- read_csv("https://grandforest.compbio.sdu.dk/files/survival_example.csv.gz")

print(D[1:8,1:8])
```

## Training the Grand Forest model

A Grand Forest model is trained on the expression data using the `grandforest` function.
We supply the expression data and network with the `data` and `graph_data` arguments. We also indicate which columns are reponse variables by setting the `dependent.variable.name` and `status.variable.name` arguments.

```{r}
library(grandforest)

model <- grandforest(
  data = D,
  graph_data = edges,
  dependent.variable.name = "os_time",
  status.variable.name = "os_event",
  num.trees=500 # 10000 trees recommended for real analysis
)

print(model)
```

In this example we only train 500 decision trees but in a real analysis we recommend using at least 10000 trees for optimal results.

Once we have trained a model we can obtain gene importance estimates using the `importance`-method. We can use this to obtain a table of the 25 most important genes:

```{r}
library(org.Hs.eg.db) # for mapping Entrez IDs to gene names

top25 <- importance(model) %>%
  sort(decreasing=TRUE) %>%
  head(25) %>%
  enframe %>%
  rename(name = "gene")  %>%
  mutate(label=mapIds(org.Hs.eg.db, gene, "SYMBOL", "ENTREZID"))

print(top25)
```

```{r, fig.width=6, fig.height=4}
ggplot(top25, aes(reorder(label, -value), value)) +
  geom_bar(stat="identity") +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  labs(x="gene", y="importance")
```

## Extract and visualize gene module

We can also visualize the gene module as a network. Here we extract the subnetwork induced by the 25 genes and visualize the network using [geomnet](https://cran.r-project.org/web/packages/geomnet/index.html).

```{r, fig.width=6, fig.height=6}
library(geomnet)

subnetwork <- filter(edges, source %in% top25$gene & target %in% top25$gene)

net.df <- fortify(as.edgedf(subnetwork), top25)

ggplot(net.df, aes(from_id=from_id, to_id=to_id)) +
  geom_net(aes(colour=importance, label=label),
    layout.alg = "circle", directed=FALSE,
    colour = "lightblue", size = 15,
    labelon = TRUE, labelcolour="black", vjust = 0.5, fontsize=3
  ) +
  theme_net()
```

## Session info

```{r echo=FALSE}
sessionInfo()
```
