---
title: "Unsupervised endophenotyping with Grand Forest"
author: "Simon J. Larsen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

This vignette demonstrates how to perform an unsupervised analysis with Grand Forest to search for *de novo* endophenotypes and gene modules in non-small cell lung cancer.

## Network preparation

First we need to prepare the gene-gene interaction network. We can obtain a protein-protein interaction network from BioGRID using the [simpIntLists](https://bioconductor.org/packages/release/data/experiment/html/simpIntLists.html) package from Bioconductor.

```{r}
library(data.table)
library(tidyverse)
library(simpIntLists)

data("HumanBioGRIDInteractionEntrezId")

# convert edge lists into two-column data frame
edges <- lapply(HumanBioGRIDInteractionEntrezId, function(x) {
  data.frame(
    source = as.character(x$name),
    target = as.character(x$interactors),
    stringsAsFactors = FALSE
  )
})
edges <- rbindlist(edges)

head(edges)
```

The resulting data frame should contain two columns with gene identifiers as `character` strings. In this example we will be using Entrez Gene IDs, but any other identifier can be used, provided the same type is used in the experimental data as well.

## Expression data preparation

Next we download a gene expression data set from non-small cell lung cancer patients.
The dataset was extracted from [GSE30219](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE30219).
The first two columns contain survival information for each patient. Since we will be doing unsupervised analysis, we will simply remove these columns.
The resulting data frame contains a column for each gene with gene expression values for each patient.

```{r}
# Download lung cancer expression data set with survival times.
D <- read_csv("https://grandforest.compbio.sdu.dk/files/survival_example.csv.gz")

# remove survival information and keep for latter
survival <- D[,1:2]
D <- D[,-(1:2)]

print(D[1:8,1:8])
```

## Training the Grand Forest model

We can train an unsupervised model using the `grandforest_unsupervised` helper function. The function will generate a background distribution from the input data and train a Grand Forest model to recognize the foreground from the background.

```{r}
library(grandforest)

model <- grandforest_unsupervised(
  data = D,
  graph_data = edges,
  num.trees = 1000
)
```

In this example we only train 1000 decision trees but in a real analysis we recommend using at least 10000 trees for optimal results.

Once we have trained a model we can obtain gene importance estimates using the `importance`-method. We can use this to obtain a table of the 25 most important genes:

```{r}
library(org.Hs.eg.db) # for mapping Entrez IDs to gene names

top25 <- importance(model) %>%
  sort(decreasing=TRUE) %>%
  head(25) %>%
  as_data_frame %>%
  rownames_to_column(var="gene") %>%
  mutate(label=mapIds(org.Hs.eg.db, gene, "SYMBOL", "ENTREZID"))

print(top25)
```

```{r, fig.width=6, fig.height=4}
ggplot(top25, aes(reorder(label, -value), value)) +
  geom_bar(stat="identity") +
  theme_classic() + theme(axis.text.x=element_text(angle=45, hjust=1)) +
  labs(x="gene", y="importance")
```

## Extract and visualize gene module

We can also visualize the gene module as a network. Here we extract the subnetwork induced by the 25 genes and visualize the network using [geomnet](https://cran.r-project.org/web/packages/geomnet/index.html).

```{r, fig.width=6, fig.height=6}
library(geomnet)

subnetwork <- filter(edges, source %in% top25$gene & target %in% top25$gene)

net.df <- fortify(as.edgedf(subnetwork), top25)

ggplot(net.df, aes(from_id=from_id, to_id=to_id)) +
  geom_net(aes(colour=importance, label=label),
    layout.alg = "circle", directed=FALSE,
    colour = "lightblue", size = 15,
    labelon = TRUE, labelcolour="black", vjust = 0.5, fontsize=3
  ) +
  theme_net()
```

## Clustering patients

We can stratify the patients into new endophenotypes by clustering them using the genes in the gene module we obtained with Grand Forest.

We first extract the 25 most important genes and scale/center the expression values. Then we cluster the patients into two groups using k-means clustering.

```{r}
# Extract module features and scale/center values.
D.scaled <- scale(D[,top25$gene], center=TRUE, scale=TRUE)
colnames(D.scaled) <- top25$label

# Cluster into two groups using k-means
cl <- kmeans(D.scaled, centers=2, nstart=20)
print(cl$cluster)
```

We can visualize the clustering as a heatmap to check which genes appear to be up- and downregulated in each group.

```{r, fig.width=6, fig.height=5}
library(ComplexHeatmap)
Heatmap(D.scaled, split = cl$cluster, name = "expression")
```

We can also use the survival information we removed earlier to compare the survival curves for the two endophenotypes we identified.
Here we observe that the stratification we identified is highly significantly associated to overall survival.

```{r, fig.width=5, fig.height=4}
library(survival)
library(survminer)
cl.survival <- data.frame(survival, cluster=cl$cluster)
ggsurvplot(survfit(Surv(os_time, os_event)~cluster, data=cl.survival), pval=TRUE)$plot
```

## Session info

```{r echo=FALSE}
sessionInfo()
```